The Markov Link Method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-15 paper/paper.tex
many notation changes

@@

\title{The Markov Link Method: a nonparametric approach to combine observations from multiple experiments}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski}

@@

The last century has seen a proliferation of experimental techniques.   Different perspectives arise from different techniques (and even from different labs performing the `same' protocol).  Ideally, this yields theories that account for all perspectives.  Instead, sometimes, this leaves us with many perspectives and no idea how they are connected.

Our goal is to connect different perspectives.  To make this idea rigorous, we assume each experiment has two aspects:
%
\begin{itemize}
    \item a \emph{sampling procedure} determines what we study
    \item a \emph{measurement procedure} defines the perspective we take in studying it.
\end{itemize}
%
If two experiments differ in both aspects, it may be difficult to gain insight by combining their data.  For example, imagine one experiment performs a transcriptomic measurement of liver cells and another applies a morphological measurement of heart cells.  Without additional experiments or knowledge, looking at these two experiments together may not be very helpful.  On the other hand, if two experiments have some aspect in common, there is something to be learned.  In this sense, a collection of experiments can be understood as a network: a pair of experiments are connected if they have some aspect in common.  In some cases, we can also connect two experiments if they differ in both aspects but we have prior knowledge (e.g.\ if the sampling procedures are different, but a reweighting can be used to correct for the difference).    When many experiments can be joined into a common network, much can be learned.

We here propose the The Markov Link Method (MLM), an algorithm that uses a network of experiments to learn the relationship between two different perspectives.  We formalize this through what we call the `measurement link' parameters, $q^*(y|x)$.  The meaning of these parameters is defined as follows.  Let us say we obtain measurement result $x$ from one measurement procedure.  The number $q^*(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.  The link tells us how the properties measured by each procedure are associated.  For example, the link lets us answer questions such as ``I have observed the transcriptome of this cell -- what morphologies might this cell have?''

If we have a single large experiment that measures each specimen with both measurement procedures, it is straightforward to estimate the link.   We can examine specimens through both perspectives simultaneously.  However, in many cases we have some experiments in which specimens are measured with one procedure, other experiments using the other procedure, and few or no experiments where specimens are measured with both procedures.  This is particularly true if the measurement techniques destroy or alter the specimen in the process of measuring it.

In this paper we focus particularly on the case that we have a network of experiments based on several sampling procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The MLM could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the MLM.

\figgy{
\hfill{}\figgygrs{.5}{../images/bigpicture}{(a) Setup: each specimen can be \\ measured with one of two procedures, \\ but can't be measured with both.\\ How do we study the relationship \\ between the procedures?}
\hfill{}\figgygrs{.4}{../images/plate}{(b) Key modeling assumption: \\ the link $q$ is the same for \\ every sampling procedure. }\hfill{}

\caption{{\bf The Markov Link Method (MLM) links different measurement procedures by using a set of different sampling procedures.}  In subfigure (a) we have neurons gathered using many different sampling procedures.  We select half of the neurons and determine what type of neuron they are, using transcriptomic measurements.  We classify the second half of neurons based on a morphological properties.  The MLM gives a way to link these two types of measurement and reconcile the two notions of cell-type.  In subfigure (b) we present a plate diagram articulating the statistical assumption that makes it possible (cf.\ \cite{koller2009probabilistic} for a thorough explanation of this type of diagram).  For each neuron in each group, this diagram considers the sampling procedure ($\ell$), the transcriptomic type ($X$), and the morphological type ($Y$).  The assumption is that the transcriptomic type is sufficiently detailed to \emph{statistically isolate} the transcriptomic type from the sampling procedure.  This is reflected in the arrows of the diagram; all paths from $\ell$ variables to $Y$ variables pass through $X$ variables.  The assumption allows us to estimate the link, $q$. \label{fig:bigpicture}}
}

@@

Imagine that we have a large collection of human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the sampling procedure used to obtain the specimen.  For example, we can select different subpopulations of cells based on proteomic markers.  For each marker we might have a different sampling procedure designed to select cells with that marker.
    \item $X$, the cell's `transcriptomic type,' as measured by looking at the transcriptomic activity in the cell.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to reconcile the two different notions of `cell-type' indicated by $X$ and $Y$.  We would like to know what transcriptomic types are associated with what morphological types.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  Our main assumption is that these kinds of associations are unaffected by the choice of sampling procedure.  In particular:

\begin{center}
\fbox{\parbox[c]{4in}{
\vspace{.1in}
\begin{center}
\textbf{The Markov Link Method Assumption}\\$\mathbb{P}(Y|X,\ell)=q^*(y|x)$ for each $\ell$
\end{center}
\vspace{.1in}
}}
\end{center}

Here $\mathbb{P}(Y|X,\ell)$ indicates the probability that the morphological type is $y$ given that the transcriptomic type is $x$ and the specimen was sampled with strategy $\ell$.  The assumption is that this probability actually does not depend upon $\ell$.  This can be understood intuitively through a thought experiment.  Let us imagine we have attained perfect understanding of cellular biology.  Someone selects a cell using one of a set of sampling procedures and measures the transcriptomic activity of the cell.  We are then told the cell's transcriptomic activity (but we are not told which of the sampling procedured was used to gather the specimen).  Using our perfect knowledge of the physical systems, we could then make predictions about a measurement of the cell's morphology.  Now we are told new information: we are told which sampling procedure was used to gather the specimen.  Would this substantially change our predictions?  If so, the MLM assumption is violated.  However, if the transcriptomic measurement is sufficiently informative, learning the sampling strategy would not substantially change our predictions about the cell morphology.  As long as we can find a measurement procedure that is sufficiently informative in this way, we can apply the MLM.  A list of examples where the MLM might apply may be found in Section \ref{sec:examples}.

Our task is to estimate the link, $q^*$, from a collection of experiments.  We suppose that for each sampling procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $N^X$ and $N^Y$.  Here $N^X_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
    \item $\hat q(y|x)$ -- a point estimate for the link parameter $q^*(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $q^*(y|x)$ with high probability
%
To produce these estimates for the link, we take a bayesian point of view.  It might seem sensible to start by specifying some prior beliefs about the link $q^*$.  However, the link is in many cases `nonidentifiable';  we describe this problem in detail in Appendix \ref{sec:casestudies}.  One consequence of nonidentifiability is that standard bayesian inferences would have no robustness to our precise choice of prior beliefs.  To avoid these problems, we focus on quantities which can be identified from the data.  In particular,
\begin{itemize}
    \item Let $p^*(x|\ell)$ denote the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$.  These probabilities can be estimated from the data, $N^X$.
    \item Let $h^*(y|\ell) \triangleq \sum_x p^*(x|\ell)q^*(y|x)$ denote the probability that a cell sampled with procedure $\ell$ will have morphological type $y$.  These probabilities can be estimated from the data, $N^Y$.
    \item Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$ denote the set of links which are consistent with $p,h$.  Notice that $q^*\in\Theta(p^*,h^*)$, but there may be other links $q\in\Theta(p^*,h^*)$.  Any such link yields the same distribution on what we can observe, $N^X,N^Y$.  Therefore there is simply no way to use data to distinguish among them.  This is the essence of the nonidentifiability problem mentioned above.  So, instead of trying to estimate this inestimable quantity, we will focus on three other quantities which can be identified from data.  Let
    \begin{itemize}
        \item $q_{\mathrm{lo},x,y} \triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x)$
        \item $q_{\mathrm{hi},x,y} \triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x)$ 
        \item $q_{\mathrm{mid}} \triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} q(y|x)^2$
    \end{itemize}
    Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known; optimization details can be found in Appendix \ref{sec:optproblems}.  Morever, even when $q^*$ is nonidentifiable, we can still be assured that $q_{\mathrm{lo}}, q_{\mathrm{hi}}$ are identifiable and $q_{\mathrm{lo},x,y} \leq q^*(y|x) \leq q_{\mathrm{hi},x,y}$.  We can also hope that $q_{\mathrm{mid}}$ will provide a single estimate which strikes a middle ground; in particular, $q_{\mathrm{mid}}$ is the link value which is closest to uniform among all links which are consistent with $p^*,h^*$.  
\end{itemize}  
To summarize: it may not be possible to estimate $q^*$ from data, but we can always estimate $p^*,h^*$.  We can also compute deterministic functions of $p^*,h^*$ which bound $q^*$, namely $q_{\mathrm{lo}},q_{\mathrm{hi}}$.  We take a Bayesian point of view in estimating $p^*,h^*$.  For prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$, namely
\[
\mathbb{P}(p^*,h^*) \propto 1
\]
Following the bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge we would like to incorporate.  First, we have observed the data, $N^X,N^Y$.  Second, we know from the MLM assumption that there exists \emph{some} value $q^*$ which is consistent with $p^*,h^*$.   To make the second idea formal, let $D(h|h')$ denote the Kullback-Leibler divergence and $G(p,h) = \min_{h':\ \Theta(p,h')\neq \emptyset} D(h|h')$.  $G(p,h)=0$ is a numerical way to indicate that the MLM assumption can hold.  After incorporating this knowledge, we find ourselevs interested in the conditional distribution
\[
\mathbb{P}\left(p^*,h^* | N^X,N^Y,G(p^*,h^*)=0\right)
\]
We use this posterior to produce estimates for our quantities of interest:
    \item $\hat q$ is calculated using posterior expectation:
    \[
    \hat q \triangleq \mathbb{E}\left[q_{\mathrm{mid}} | N^X,N^Y,G(p^*,h^*)=0\right]
    \]
    %
    \item $C_{x,y}$ is calculated using credible intervals.  For each $x,y$, we define $C_{x,y}$ to be the largest interval such that 
    \begin{align*}
    \mathbb{P}\left(q_{\mathrm{lo},x,y} \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% \\
    \mathbb{P}\left(q_{\mathrm{hi},x,y} \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% 
    \end{align*}
Code to compute these objects is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

In practice, we were not able to find a way to compute the credible interval and posterior means exactly.  It seems difficult to even sample from the conditional distribution.  Common approaches to this type of problem involve Markov-Chain Monte Carlo and Variational Inference, but we were unable to make these approaches work in practice.  It seems nontrivial to work with the condition $G(p,h)=0$ that formalizes the MLM assumption.  We instead take a somewhat na\"ive approach.  We start by defining $\tilde P,\tilde H$ to carry the conditional distribution of $p^*,h^*$ given the observed data (but not conditioned on the fact that the MLM assumption holds).  We then simply project $\tilde H$ to the space of parameters which are consistent with the MLM assumption.  That is, we take $\hat H$ to be the solution of the minimization problem $\min_{h'}D(\tilde H|h')$, subject to the constraint that $E(\tilde P,h')=0$.  Optimization details can be found in Appendix \ref{sec:optproblems}.  We hope that the distribution of $\tilde P,\hat H$ will approximate the desired posterior.  We then use samples from this distribution to compute what we need.  We asymptotically expect that $\tilde P,\tilde H$ will nearly satisfy the MLM assumption in any case, so this approximation should not make a large difference.  However, we caution that the exact implications of the approximation are difficult to reason about.   In any case, it is a practical solution to a difficult problem.

@@

\figgy{
\hfill{}\figgygrs{.4}{../images/simulationI}{(a) Estimation error, \\ varying number of samples \\ and number of sampling procedures.}\hfill{}
\hfill{}\figgygrs{.5}{../images/simulationIconf}{(b) Credible intervals, \\ varying number of samples \\ and number of sampling procedures.}\hfill{}

@@

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, sampling procedures, and the link parameters.  In each trial our `ground truth' link parameters picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct sampling procedures.  In subfigure (b) we consider the MLM's credible intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link parameter $q^*(y|x)$ with the MLM interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link parameters lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The intervals mostly contain the truth, and get narrower with more samples as long as there are enough distinct sampling procedures.   In subfigure (c) we focus on the case of exactly $4$ sampling procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is independent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of sampling procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

There are three desiderata we might hope the Markov Link Method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat q(y|x)$ should converge to the true link $q^*(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $q^*(y|x)$.
    \item[Conservative coverage] $q^*(y|x) \in C_{x,y}$ with high probability.
\end{description}
However, due to certain identifiability issues, it is only actually possible to guarantee the last desideratum.  In many cases, the MLM achieves all three.  However, we can show that identifiability issues block any possibility of success in some cases; we detail this problem in Appendix \ref{sec:casestudies}.  The good news is that even though the Markov Link Method credible intervals may not concentrate around the truth, they do as well as possible.  In particular, they do appear to converge around the quantities we can estimate, namely the $q_\mathrm{lo},q_\mathrm{hi}$ bounds described in Section \ref{sec:mlm}.  A precise  statement of what we have been able to prove may be found in Theorem \ref{thm:mainthm} of Appendix \ref{sec:casestudies}.

@@

\begin{itemize}
    \item How does MLM performance depend on the number of samples and the number of sampling procedures?  We suppose we have two measurement procedures which return categorical measurements among six categories.  Thus the first procedure yields $X\in \{1,2,3,4,5,6\}$ and the second yields $Y\in \{1,2,3,4,5,6\}$.  To see how the method performs in different circumstances, we will run many trials.  In each trial we fix the number of sampling procedures and pick `ground truth' parameters uniformly at random from the parameter space.  We then simulate a dataset from these ground truth parameters, fixing the total number of samples (spreading these samples equally among all combinations of sampling procedure and measurement procedure).  Finally, we apply the MLM to the simulated data to get the point estimates $\tilde q(y|x)$ and the credible intervals $C_{x y}$.  We measure the overall estimator convergence using a kind of total variation distance:
    \[
    \mathrm{Error}(\tilde q,q^*)=\frac{1}{12}\sum_{x=1}^{6}\sum_{y=1}^{6} |\tilde q(y|x) - q^*(y|x)|
    \]
    This error ranges between zero and one.  Zero error indicates that $\tilde q=q^*$. An error of one indicates that the estimate has completely incorrect beliefs about the probability mass, i.e. $q^*(y|x)=0$ whenever $\tilde q(y|x)>0$ and vice-versa.  We also look at the MLM credible intervals and see how often they cover the true parameters.

    The results are summarized in Figure \ref{fig:simI}.  In trials with more samples, the estimator generally has lower error.  However, to make the error actually converge to zero we need at least six distinct sampling procedures.  This figure also shows that the intervals work correctly regardless of of the number of samples or sampling procedures; they include the ground truth with high probability.   With many samples and sampling procedures, the intervals are small and concentrated around the truth.  With fewer samples or sampling procedures, the intervals are sometimes forced to be larger.  However, our uncertainty is not the same for every parameter.  In some cases we are able to obtain a very tight credible interval for one particular parameter even though the overall estimator error is high.

    \item How does estimator convergence depend upon the link parameters themselves?  In each trial of this simulation we use four sampling procedures and our measurement procedures yield one of six categories.  In the previous simulations we saw that estimator convergence was generally impossible in this situation, due to the small number of sampling procedures.  However, in our previous simulations we picked the link parameters uniformly from parameter space.  Now we will be more choosey.   On one extreme, we will produce trials where the link makes $X,Y$ independent, i.e. $q^*(y|x)=q^*(y|x')$ for every $x,x'$.  On the other extreme, we will have trials where $X,Y$ are deterministically related by the equation $X=Y$.  We will also consider every link ``in-between'' these two extremes (found by convex combinations).  Figure \ref{fig:simI} shows that estimator convergence is actually possible in the two extreme cases.  However, estimator convergence fails for the in-between cases.  We leave a complete mathematical understanding of this result to future work.  For the purposes of this paper, we content ourselves by noting one interesting case: let the measurements return one of $2^k$ categories and let us use only $k+1$ carefully chosen sampling procedures.  Now suppose the link defines any invertible deterministic function between $X$ and $Y$.  In this case, with enough enough samples, we can determine both that the relationship is deterministic and the exact specification of the invertible function.  This result is proven in Appendix \ref{sec:casestudies}, Theorem \ref{thm:miracle}.
\end{itemize}

@@

>>>>

\newcommand{\plx}{\ensuremath{f}}
\newcommand{\pxy}{\ensuremath{g}}
\newcommand{\ply}{\ensuremath{h}}
% \newcommand{\DX}{\ensuremath{{\mathcal{D}^X}}}
% \newcommand{\DY}{\ensuremath{{\mathcal{D}^Y}}}
\newcommand{\DX}{\ensuremath{\mathcal{D}^X}}
\newcommand{\DY}{\ensuremath{\mathcal{D}^Y}}
\newcommand{\MLMH}{\ensuremath{\mathcal{A}}}

\newcommand{\MLM}{MLM}

@@

\title{The Markov Link Method: a nonparametric approach to combine observations from multiple experiments}
\author{Jackson Loper, Osnat Penn, Trygve Bakken, David Blei, Liam Paninski, \\Additional Authors To Be Determined}

@@

Imagine we have a collection of different experiments about human cells.  We would like to link the different perspectives that we get from each experiment.  To make this idea rigorous, we need to define what we mean by `perspective' and what we mean by `link.'

To define `perspective,' we assume each experiment has two aspects:
%
\begin{itemize}
    \item A \emph{selection procedure} defines how we select cells.  For example, one experiment might select only liver cells, and another might select only cells which express the Vip protein.  
    \item A \emph{measurement procedure} defines what we measure about each cell.  For example, one experiment might measure the transcriptomic profile of a cell, and another experiment might measure the cell morphology.
\end{itemize}
%
The selection procedure defines what the experiment studies, and the measurement procedures determines the experiment's `perspective.'

The `link' between two measurement procedures should enable us to answer questions such as ``I have observed the transcriptome of this cell -- what morphologies might this cell have?''  We formalize this idea through what we call the `measurement link parameters,' $\pxy(y|x)$.  Specifically, let us say we obtain measurement result $x$ from one measurement procedure.  The number $\pxy(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.

We here propose the The Markov Link Method (\MLM), an algorithm that uses a collection of experiments to estimate the measurement link parameters, \pxy.  In this paper we focus particularly on the case that we have experiments based on several selection procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The \MLM{} could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the \MLM.  It embodies a common situation in biological problems, such as is found in \cite{gouwens2018classification}.

\figgy{
\hfill{}\figgygrs{.5}{../images/bigpicture}{(a) Setup: each specimen can be \\ measured with one of two procedures, \\ but can't be measured with both.\\ How do we study the relationship \\ between the procedures?}
\hfill{}\figgygrs{.4}{../images/plate}{(b) Key modeling assumption: \\ the link $\pxy$ is the same for \\ every selection procedure. }\hfill{}

\caption{{\bf The Markov Link Method (MLM) links different measurement procedures by using a set of different selection procedures.}  In subfigure (a) we have neurons gathered using many different selection procedures.  We select half of the neurons and determine what type of neuron they are, using transcriptomic measurements.  We classify the second half of neurons based on a morphological properties.  The MLM gives a way to link these two types of measurement and reconcile the two notions of cell-type.  In subfigure (b) we present a plate diagram articulating the statistical assumption that makes it possible (cf.\ \cite{koller2009probabilistic} for a thorough explanation of this type of diagram).  For each neuron in each group, this diagram considers the selection procedure ($\ell$), the transcriptomic type ($X$), and the morphological type ($Y$).  The assumption is that the transcriptomic type is sufficiently detailed to \emph{statistically isolate} the transcriptomic type from the selection procedure.  This is reflected in the arrows of the diagram; all paths from $\ell$ variables to $Y$ variables pass through $X$ variables.  The assumption allows us to estimate the link, $\pxy$. \label{fig:bigpicture}}
}

@@

The Markov Link Method is a general way to estimate measurement link parameters from a collection of experiments.  For concreteness, we will use experiments about human cells as a running example.  Let us say that we have gathered many human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the selection procedure used to obtain the specimen
    \item $X$, the cell's `transcriptomic type,' as measured through a transcriptomic analysis.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to link the transcriptomic and morphological perspectives on `cell-type,' as manifested through the $X$ and $Y$ measurement procedures.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  Our main assumption is that these kinds of associations are unaffected by the choice of selection procedure.  In particular:

\begin{center}
\fbox{\parbox[c]{4in}{
\vspace{.1in}
\begin{center}
\textbf{The Markov Link Method Assumption}\\$\mathbb{P}(Y|X,\ell)=\pxy(y|x)$ for each $\ell$
\end{center}
\vspace{.1in}
}}
\end{center}

Here $\mathbb{P}(Y|X,\ell)$ indicates the probability that the morphological type is $y$ given that the transcriptomic type is $x$ and the specimen was sampled with strategy $\ell$.  The assumption states that that this probability actually does not depend upon $\ell$.  This assumption can be understood intuitively through a thought experiment.  Let us imagine we have attained perfect understanding of cellular biology.  Someone selects a cell using one of a set of selection procedures and measures the transcriptomic activity of the cell.  We are then told the cell's transcriptomic type (but we are not told which of the selection procedures was used to gather the specimen).  Using our perfect knowledge of the physical systems, we could then make predictions about the cell's morphological type.  Now we are told new information: we are told which selection procedure was used to gather the specimen.  Would this substantially change our predictions?  If so, the MLM assumption is violated.  However, if the transcriptomic measurement is sufficiently informative, learning the sampling strategy would not substantially change our predictions about the cell morphology.  As long as we can find a measurement procedure that is sufficiently informative in this way, we can apply the MLM.  A list of examples where the MLM might apply may be found in Section \ref{sec:examples}.

More formally, our task is to estimate the link parameters $\pxy$ from a collection of experiments.  We suppose that for each selection procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $\DX$ and $\DY$.  Here $\DX_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $\DY_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
\begin{enumerate}
    \item $\hat \pxy(y|x)$ -- a point estimate for the link parameter $\pxy(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $\pxy(y|x)$ with high probability
\end{enumerate}
%
To estimate these, we must consider certain `nuisance parameters.'  Our observable data, $\DX,\DY$, is governed by the following parameters:
\begin{itemize}
    \item $\plx(x|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$
    \item $\ply(y|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have morphological type $y$
\end{itemize}
It is straightforward to estimate $\plx,\ply$ from $\DX,\DY$ using standard methods.  We would like to use these parameters to tell us something about our object of interest, the link $\pxy$.  Under the Markov Link Method assumption, this link is connected to $\plx,\ply$ through the equations
\begin{gather}\label{eq:mlmh}
\sum_x \plx(x|\ell)\pxy(y|x)=\ply(y|\ell) \qquad \forall y,\ell
\end{gather}
With these equations and knowledge of $\plx,\ply$, we can produce an estimate for the link: $\hat \pxy_{\plx,\ply}$.  We can also produce bounds for the link.  We define these objects precisely in Appendix \ref{sec:mlmdetails}.  However, in practice we do not actually have exact knowledge of $\plx,\ply$.  We only have estimates from data.  We take a Bayesian perspective to account for our uncertainty in this estimation.  For our prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$.  Following the Bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge about $\plx,\ply$.  First, we have observed the data, $\DX,\DY$.  Second, we know from the MLM assumption that there exists \emph{some} value $\pxy$ such that Equation (\ref{eq:mlmh}) holds; let \MLMH{} denote the event that this holds.  We then define our point estimate $\hat \pxy$ as the posterior expectation of $\hat \pxy_{\plx,\ply}$ after conditioning on this knowledge, i.e.
\[
\hat \pxy(y|x) = \mathbb{E}\left[\hat \pxy_{\plx,\ply} | \DX,\DY,\MLMH \right]
\]
We can use a similar strategy to produce the credible intervals $C_{x,y}$.  Exact details can be found in Appendix \ref{sec:mlmdetails}.  Code to compute $\hat \pxy(y|x)$ and $C_{x,y}$ is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

@@

\figgy{
\hfill{}\figgygrs{.4}{../images/simulationI}{(a) Estimation error, \\ varying number of samples \\ and number of selection procedures.}\hfill{}
\hfill{}\figgygrs{.5}{../images/simulationIconf}{(b) Credible intervals, \\ varying number of samples \\ and number of selection procedures.}\hfill{}

@@

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, selection procedures, and the link parameters.  In each trial our `ground truth' link parameters picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct selection procedures.  In subfigure (b) we consider the MLM's credible intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link parameter $\pxy(y|x)$ with the MLM interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link parameters lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The intervals mostly contain the truth, and get narrower with more samples as long as there are enough distinct selection procedures.   In subfigure (c) we focus on the case of exactly $4$ selection procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is independent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of selection procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

There are three desiderata we might hope the Markov Link Method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat \pxy(y|x)$ should converge to the true link $\pxy(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $\pxy(y|x)$.
    \item[Conservative coverage] $\pxy(y|x) \in C_{x,y}$ with high probability.
\end{description}
However, due to certain identifiability issues, it is only actually possible to guarantee the last desideratum.  In many cases, the MLM achieves all three.  However, we can show that identifiability issues block any possibility of success in some cases; we detail this problem in Appendix \ref{sec:casestudies}.  The good news is that even though the Markov Link Method credible intervals may not concentrate around the truth, they do as well as possible.  In particular, they do appear to converge around the quantities we can estimate, namely the $q_\mathrm{lo},q_\mathrm{hi}$ bounds described in Appendix \ref{sec:mlmdetails}.  A precise statement of what we have been able to prove may be found in Theorem \ref{thm:mainthm} of Appendix \ref{sec:casestudies}.

@@

\begin{itemize}
    \item How does MLM performance depend on the number of samples and the number of selection procedures?  We suppose we have two measurement procedures which return categorical measurements among six categories.  Thus the first procedure yields $X\in \{1,2,3,4,5,6\}$ and the second yields $Y\in \{1,2,3,4,5,6\}$.  To see how the method performs in different circumstances, we will run many trials.  In each trial we fix the number of selection procedures and pick `ground truth' parameters uniformly at random from the parameter space.  We then simulate a dataset from these ground truth parameters, fixing the total number of samples (spreading these samples equally among all combinations of selection procedure and measurement procedure).  Finally, we apply the MLM to the simulated data to get the point estimates $\tilde q(y|x)$ and the credible intervals $C_{x y}$.  We measure the overall estimator convergence using a kind of total variation distance:
    \[
    \mathrm{Error}(\tilde q,\pxy)=\frac{1}{12}\sum_{x=1}^{6}\sum_{y=1}^{6} |\tilde q(y|x) - \pxy(y|x)|
    \]
    This error ranges between zero and one.  Zero error indicates that $\tilde q=\pxy$. An error of one indicates that the estimate has completely incorrect beliefs about the probability mass, i.e. $\pxy(y|x)=0$ whenever $\tilde q(y|x)>0$ and vice-versa.  We also look at the MLM credible intervals and see how often they cover the true parameters.

    The results are summarized in Figure \ref{fig:simI}.  In trials with more samples, the estimator generally has lower error.  However, to make the error actually converge to zero we need at least six distinct selection procedures.  This figure also shows that the intervals work correctly regardless of of the number of samples or selection procedures; they include the ground truth with high probability.   With many samples and selection procedures, the intervals are small and concentrated around the truth.  With fewer samples or selection procedures, the intervals are sometimes forced to be larger.  However, our uncertainty is not the same for every parameter.  In some cases we are able to obtain a very tight credible interval for one particular parameter even though the overall estimator error is high.

    \item How does estimator convergence depend upon the link parameters themselves?  In each trial of this simulation we use four selection procedures and our measurement procedures yield one of six categories.  In the previous simulations we saw that estimator convergence was generally impossible in this situation, due to the small number of selection procedures.  However, in our previous simulations we picked the link parameters uniformly from parameter space.  Now we will be more choosey.   On one extreme, we will produce trials where the link makes $X,Y$ independent, i.e. $\pxy(y|x)=\pxy(y|x')$ for every $x,x'$.  On the other extreme, we will have trials where $X,Y$ are deterministically related by the equation $X=Y$.  We will also consider every link ``in-between'' these two extremes (found by convex combinations).  Figure \ref{fig:simI} shows that estimator convergence is actually possible in the two extreme cases.  However, estimator convergence fails for the in-between cases.  We leave a complete mathematical understanding of this result to future work.  For the purposes of this paper, we content ourselves by noting one interesting case: let the measurements return one of $2^k$ categories and let us use only $k+1$ carefully chosen selection procedures.  Now suppose the link defines any invertible deterministic function between $X$ and $Y$.  In this case, with enough enough samples, we can determine both that the relationship is deterministic and the exact specification of the invertible function.  This result is proven in Appendix \ref{sec:casestudies}, Theorem \ref{thm:miracle}.
\end{itemize}

@@

\section{Exact details of the Markov Link Method}

\label{sec:mlmdetails}
